Staged Abstract Interpreters

### 2018-11-17 paper/10_intro.tex
revise

It has been observed that partially applying context-sensitivity on selected portion of the program could
improve the precision and efficiency \cite{zipper2018, Kastrinis:2013:HCP:2491956.2462191}.
We show that staging abstract interpreters as an approach to effectively implement hybird context-sensitivity\todo{}.

@@

\begin{itemize}
  \item Starting from a generic definition of interpreter for a higher-order functional language,
    whose syntax and semantics are described in Section~\ref{bg_lang}, we show its concrete
    interpreter (Section~\ref{unstaged_conc}), staged concrete interpreter (Section~\ref{stagedinterp}), 
    abstract interpreter (Section~\ref{unstaged_abs}) and staged abstract interpreter (Section~\ref{sai}).
    Our design allows one can derive these four different kinds of interpreters by mechanically changing
    the underlying semantics or stage annotations.
    %We later extend the staged abstract interpreter to support imperative language features (Section~\ref{cases_imp}), such as states and loops.
  \item We demonstrate that the staged abstract interpreter not only improves the \textit{efficiency}
    but also the \textit{scalability}, which are two major issues in static analysis, if we apply it on open
    programs. \todo{Section X}.
  \item To demonstrate the applicability and usefulness of the proposed approach, we conduct three case studies 
    Section~\ref{cases_study}: 
    \subitem (1) We revisit Abstract Compilation \cite{Boucher:1996:ACN:647473.727587} in Section~\ref{cs_ac}, 
    which was originally a closure generation technique applied on 0-CFA.
    We show that abstract compilation can be understood and easily implemented as an instance of our staged abstract interpreter.
    But unlike the abstract compilation, our approach does not require the implementers to rewrite the analysis.
    \subitem (2) We demonstrate pushdown control-flow analysis techniques (Section~\ref{cfa}), such as context-sensitive 
    analysis and more precise stores can be integrated to staged abstract interpreters without losing any precision.
    \subitem (3) We show that in an imperative language (Section~\ref{cases_imp}), by replacing the abstract domain
    to an interval domain (as well as staged), the staged abstract interpreter performs numerical analysis.
    % (Optional: taint/information flow analysis?)
  \item We empirically evaluate the performance improvement by staging an big-step abstract interpreter 
    for control-flow analysis (Section~\ref{evaluation}). 
    We compare our staged abstract interpreter with the unstaged version.
\end{itemize}

>>>>

\iffalse
It has been observed that partially applying context-sensitivity on selected portion of the program could
improve the precision and efficiency \cite{zipper2018, Kastrinis:2013:HCP:2491956.2462191}.
We show that staging abstract interpreters as an approach to effectively implement hybird context-sensitivity\todo{}.
\fi

@@

\begin{itemize}
  \item Starting from a generic definition of interpreter for a higher-order functional language,
    whose syntax and semantics are described in Section~\ref{bg_lang}, we show its concrete
    interpreter (Section~\ref{unstaged_conc}), staged concrete interpreter (Section~\ref{stagedinterp}), 
    abstract interpreter (Section~\ref{unstaged_abs}) and staged abstract interpreter (Section~\ref{sai}).
    Our design allows one can derive these four different kinds of interpreters by mechanically changing
    the underlying semantics or stage annotations.
    %We later extend the staged abstract interpreter to support imperative language features (Section~\ref{cases_imp}), such as states and loops.
  \item As shown in Section~\ref{sai}, we demonstrate that if we apply the abstract interpreters on open 
    programs, it not only improves the \textit{efficiency} but also the \textit{scalability}, 
    which are two major issues in static analysis.
  \item To demonstrate the applicability and usefulness of the proposed approach, we conduct three case studies 
    Section~\ref{cases_study}: 
    \subitem (1) We revisit Abstract Compilation \cite{Boucher:1996:ACN:647473.727587} in Section~\ref{cs_ac}, 
    which was originally a closure generation technique applied on 0-CFA.
    We show that abstract compilation can be understood and easily implemented as an instance of our staged abstract interpreter.
    But unlike the abstract compilation, our approach does not require the implementers to rewrite the analysis.
    \subitem (2) We demonstrate pushdown control-flow analysis techniques (Section~\ref{cfa}), such as context-sensitive 
    analysis and more precise stores can be integrated to staged abstract interpreters without losing any precision.
    \subitem (3) We show that in an imperative language (Section~\ref{cases_imp}), by replacing the abstract domain
    to an interval domain (as well as staged), the staged abstract interpreter performs numerical analysis.
    % (Optional: taint/information flow analysis?)
  \item We empirically evaluate the performance improvement by staging an big-step abstract interpreter 
    for control-flow analysis (Section~\ref{evaluation}). 
    We compare our staged abstract interpreter with the unstaged version.
\end{itemize}

### 2018-11-17 paper/30_sai.tex
revise

Motivation: one of the challenges of modern static analysis is program usually depends on
large libraries programs\cite{toman_et_al:LIPIcs:2017:7121}. 
Can we analyze programs and libraries separately without losing precision? So that we can 
reduce part of the overhead of repeatedly analyzing libraries for different programs.
Similarly, some static analyzers compute summary for a function or a module, that can be reused
later (like Facebook Infer). But to my knowledge, they are mostly too conservative (context-insensitive) 
or unsound, which both lead to imprecision.

Application: for example, k-CFA (k > 0) is naturally a kind of whole program analysis,
because it is interprocedural and need the last k calling contexts to distinguish
different call sites.
But can we analyze programs (libraries) separately which generate the specialized 
analysis and leave the unavailable programs (for the moment) as dynamic parameters, 
and then install these contexts when we have the whole program.

Another perspective: programs are data for an abstract interpreter, so if we have $n$ programs, 
then maybe there can be $n$ stages. 
Probably we can analyze first $m$ programs, and generate a residual abstract interpreter
waiting for the rest $(n-m)$ programs.
These $(n-m)$ programs might be (abstract) arguments for the first $n$ programs, and
the abstract interpreter itself might be a partial abstract interpreter.

@@

\paragraph{Big-step vs Small-step.}
What we implemented is a big-step, compositional \todo{or semi-compositional} abstract interpreter, 
where "compositional" means that every recursive call of our abstract interpreter is applied to proper 
substructures of the current syntactic parameters \cite{10.1007/3-540-61580-6_11}.
This compositionality ensures that specialization can be done by unfolding,
as well as that the specialization procedure terminates.
Nevertheless, it is also possible to specialize a small-step operational abstract semantics ---
\citeauthor{Johnson:2013:OAA:2500365.2500604}
showed this in abstract compilation \cite{Boucher:1996:ACN:647473.727587} style as one of their optimizations
of Abstract Abstract Machines \cite{Johnson:2013:OAA:2500365.2500604}. 
However, the generated abstract bytecode still requires another small-step abstract machine 
to execute, which is an additional engineering efforts.
Another alternative approach for efficient specialization is to write the abstract interpreter 
in a big-step, monadic style \cite{DBLP:journals/pacmpl/DaraisLNH17}.

>>>>

One of the challenges of modern static analysis is program usually depends on
large libraries programs~\cite{toman_et_al:LIPIcs:2017:7121}. 
Can we analyze programs and libraries separately and reuse the result without losing precision? 
Then we can reduce part of the overhead of repeatedly analyzing libraries for different programs.
Indeed, some static analyzers compute summary for a function or a module, which can be reused
later, however they are mostly too conservative or unsound, which both lead to imprecision.

The specialization of abstract interpreter provides a chance to obtain such partial analysis 
result in a mechanized way, but still keeps the analysis sound.
As we see when compiling the closures, we can specialize the abstract interpreter
with respect the body expression of the lambda term without knowing the actual argument.
The programs with some unknown variables are open programs, which is exactly the case 
if we want to analyze programs in a modular way.

For a concrete analysis, for example, $k$-CFA ($k$ > 0) is naturally a whole-program analysis,
because it is inter-procedural and needs the last $k$ calling contexts to distinguish
different call sites, where the calling contexts are dynamic values.
However, it is possible to analyze programs (libraries) separately through specializing an abstract interpreter 
that generates the analysis as the next-stage program and leave the unavailable programs and calling contexts 
as dynamic parameters, and then install these contexts when we have the whole program.

\iffalse
Another perspective: programs are data for an abstract interpreter, so if we have $n$ programs, 
then maybe there can be $n$ stages. 
Probably we can analyze first $m$ programs, and generate a residual abstract interpreter
waiting for the rest $(n-m)$ programs.
These $(n-m)$ programs might be (abstract) arguments for the first $n$ programs, and
the abstract interpreter itself might be a partial abstract interpreter.
\fi

@@

\paragraph{Big-step vs Small-step.}
What we implemented is a big-step, compositional abstract interpreter, 
where "compositional" means that every recursive call of our abstract interpreter is applied to proper 
substructures of the current syntactic parameters \cite{10.1007/3-540-61580-6_11}.
This compositionality ensures that specialization can be done by unfolding,
as well as that the specialization procedure terminates.
Nevertheless, it is also possible to specialize a small-step operational abstract semantics ---
\citeauthor{Johnson:2013:OAA:2500365.2500604}
showed this in abstract compilation \cite{Boucher:1996:ACN:647473.727587} style as one of their optimizations
of Abstract Abstract Machines \cite{Johnson:2013:OAA:2500365.2500604}. 
However, the generated abstract bytecode still requires another small-step abstract machine 
to execute, which is an additional engineering efforts.
Another alternative approach for efficient specialization is to write the abstract interpreter 
in a big-step, monadic style \cite{DBLP:journals/pacmpl/DaraisLNH17}.
